Designing Laboratory Notebooks for Managing Software-based Scientific Studies

### 2023-09-05 Dhruv-Dubey-LabNotebooks/Paper.tex
remove comments and highlighting

\begin{abstract}
%
Scientific processes rely on software as an important tool for data acquisition, analysis, and discovery. Over the years sustainable software development practices have made progress in being considered as an integral component of research. However, management of computation-based scientific studies is often left to individual researchers who design their computational experiments based on personal preferences and the nature of the study. 
%Although applications are available to create reproducible capsules for data generation, they do not provide tools to manage research in a structured way to enhance knowledge related to decisions made by researchers to configure their software instruments. 
We believe that the quality, efficiency, and reproducibility of computation-based scientific research can be improved by explicitly creating an execution environment that allows researchers to provide a clear record of traceability. This is particularly relevant to complex computational studies in high-performance computing (HPC) environments.
%Such an environment, which is analogous to a laboratory space in experimental sciences, should include the integration of a lab notebook that enables systematic curation of the research process and documents software configuration, motivation for decisions, and other options. 
In this article, we review the documentation required to maintain a comprehensive record of  HPC computational experiments for reproducibility. We also provide an overview of tools and practices that we have developed to perform such 
%different computational fluid dynamics, optimization, and performance 
studies around Flash-X, a multi-physics scientific software.
%, and demonstrate how the research process can be efficiently organized without having to compromise researcher preferences or requirements. The lightweight tools developed for the study can be applied to computational experiments that rely on a Unix-style interface for development and execution.

@@

\section{LABORATORY NOTEBOOKS FOR COMPUTATIONAL EXPERIMENTS}
Since computational work often involves coding, simulations, and data
analysis using software tools, the lab notebook is best maintained in
electronic format, and could include a combination of repositories,
spreadsheets, text, and markdown formats. 
For reasons discussed previously\footnote{\url{https://www.youtube.com/watch?v=fWpI4S_dvhc}}, the use of some form of
lab notebook in computational science has several challenges.
However, fundamental sets of activities that must be recorded are fairly
simple to enumerate.

@@

An example of a workflow that might be encountered by a developer who is also a user of the code for doing scientific investigations is shown in Figure \ref{fig:case-study-example}a. Such developers may perform regular testing of the physics and infrastructure components, keep track of the performance changes, and conduct scientific experiments at the same time.  The collection of lab notebooks \cite{Lab-Notebooks} described in Figure \ref{fig:case-study-example}b provides a good example of what may be done in such circumstances. These notebooks are all seeded from a general notebook \cite{Multiphysics-Simulations} being maintained by the developer in question. In this instance, the developer is simultaneously conducting the following experiments,

@@

\begin{figure}[h]
\centerline{\includegraphics[width=18.5pc]{case-study.png}}
\caption{(a) A common scenario in computational sciences where a developer has to conduct individual experiments and collaborate with other researchers who may not be experts in using their software. (b) Example use case of lab notebooks for Flash-X development. The developer maintains a personal repository that serves as a seed for setting up experiments for physics and performance related studies that can be shared with collaborators.}%\vspace*{-5pt}
\label{fig:case-study-example}
\end{figure}

@@

The process of seeding and managing these notebooks is implemented using a lightweight command line tool, Jobrunner \cite{Jobrunner}, that enforces a directory-based inheritance for application configurations encoded in shell scripts to manage computational experiments. The directory tree is designed to split different shell commands that set up the software stack and simulation in an optimal way. For instance, a combination of {environment.sh} and {setupAMReX.sh} is used to configure AMReX most suitably for the experiment. Here, {environment.sh} documents and sets variables that store compiler information and installation location of different libraries, and {setupAMReX.sh} stores information specific to building AMReX with the necessary options. The {Jobfiles} that exist at every level of the tree encode information that Jobrunner can use to stitch together these bash files to perform different tasks. 
%
\begin{figure}
\begin{minted}[
    gobble=1,
    frame=single,
  ]{yaml}
  # Location:
  # Experiment
  job:
    setup:
        - environment.sh
    submit:
        - environment.sh

@@

Jobrunner hides error-prone individual steps of the experiments while allowing organized and explicit documentation of configuration options that can be easily modified by editing the shell scripts. The directory tree in Figure \ref{fig:dirtree} can be easily modified and redesigned based on the requirements of the experiments to include markdown notes and analysis files. For full functionality and documentation for Jobrunner see A. Dhruv \cite{Jobrunner}.
%
\begin{figure}
\begin{minted}[
    gobble=8,
    frame=single,
  ]{bash}
        # Setting up software and dependencies
        jobrunner setup software/amrex
        jobrunner setup software/flashx

@@

%\bibliographystyle{abbrv}
%\nocite{*}
%\bibliography{References/references}
%

%\newpage
\begin{IEEEbiography}{Akash Dhruv,} is a Postdoctoral Fellow in the Mathematics and Computer Science Division at Argonne National Laboratory where he performs research in DevOps and performance engineering for scientific computing applications, along with building computational pipelines to integrate simulation with machine learning workflows. He received his Ph.D. in mechanical and aerospace engineering from George Washington University, Washington D.C., and a B.Tech in mechanical engineering from National Institute of Technology, Surat. \vspace*{8pt}
\end{IEEEbiography}

%\begin{IEEEbiography}{Jared O'Neal,} {\,} is a Software Engineer.
%\end{IEEEbiography}

\begin{IEEEbiography}{Anshu Dubey,} is a Senior Computational Scientist in the Mathematics and Computer Science Division at Argonne National Laboratory and a Senior Scientist
(CASE) at the University of Chicago. She was previously at Lawrence Berkeley National Laboratory, and before that she was the associate director of the Flash Center for Computational Science at the University of Chicago and also led the CS/Applications
group, who develop, maintain, and distribute the FLASH code. She
received her Ph.D. in computer science from Old Dominion University and
a B.Tech. in electrical engineering from the Indian Institute of Technology,
New Delhi.\vspace*{8pt}%\vadjust{\vfill\pagebreak}
\end{IEEEbiography}

>>>>

\begin{abstract}
%
Scientific processes rely on software as an important tool for data acquisition, analysis, and discovery. Over the years sustainable software development practices have made progress in being considered as an integral component of research. However, management of computation-based scientific studies is often left to individual researchers who design their computational experiments based on personal preferences and the nature of the study. We believe that the quality, efficiency, and reproducibility of computation-based scientific research can be improved by explicitly creating an execution environment that allows researchers to provide a clear record of traceability. This is particularly relevant to complex computational studies in high-performance computing (HPC) environments. In this article, we review the documentation required to maintain a comprehensive record of  HPC computational experiments for reproducibility. We also provide an overview of tools and practices that we have developed to perform such studies around Flash-X, a multi-physics scientific software.

@@

\section{LABORATORY NOTEBOOKS FOR COMPUTATIONAL EXPERIMENTS}
Since computational work often involves coding, simulations, and data
analysis using software tools, the lab notebook is best maintained in
electronic format, and could include a combination of repositories,
spreadsheets, text, and markdown formats. 
{For reasons discussed previously\footnote{\url{https://www.youtube.com/watch?v=fWpI4S_dvhc}}}, the use of some form of
lab notebook in computational science has several challenges.
{However, fundamental sets of activities that must be recorded are fairly
simple to enumerate}.

@@

An example of a workflow that might be encountered by a developer who is also a user of the code for doing scientific investigations is shown in Figure \ref{fig:case-study-example}a. Such developers may perform regular testing of the physics and infrastructure components, keep track of the performance changes, and conduct scientific experiments at the same time.  {The collection of lab notebooks \cite{Lab-Notebooks} described in Figure \ref{fig:case-study-example}b provides a good example of what may be done in such circumstances}. These notebooks are all seeded from a general notebook \cite{Multiphysics-Simulations} being maintained by the developer in question. In this instance, the developer is simultaneously conducting the following experiments,

@@

\begin{figure}[h]
\centerline{\includegraphics[width=18.5pc]{case-study.png}}
\caption{(a) A common scenario in computational sciences where a developer has to conduct individual experiments and collaborate with other researchers who may not be experts in using their software. (b) Example use case of lab notebooks for Flash-X development. The developer maintains a personal repository that serves as a seed for setting up experiments for physics and performance related studies that can be shared with collaborators.}
\label{fig:case-study-example}
\end{figure}

@@

The process of seeding and managing these notebooks is implemented using a lightweight command line tool, Jobrunner \cite{Jobrunner}, that enforces a directory-based inheritance for application configurations encoded in shell scripts to manage computational experiments. The directory tree is designed to split different shell commands that set up the software stack and simulation in an optimal way. For instance, a combination of {environment.sh} and {setupAMReX.sh} is used to configure AMReX most suitably for the experiment. Here, {environment.sh} documents and sets variables that store compiler information and installation location of different libraries, and {setupAMReX.sh} stores information specific to building AMReX with the necessary options. { The {Jobfiles} that exist at every level of the tree encode information that Jobrunner can use to stitch together these bash files to perform different tasks.} 
%
\begin{figure}
\begin{minted}[
    gobble=1,
    frame=single,
  ]{yaml}
  # Location:
  # Experiment
  job:
    setup:
        - environment.sh
    submit:
        - environment.sh

@@

Jobrunner hides error-prone individual steps of the experiments while allowing organized and explicit documentation of configuration options that can be easily modified by editing the shell scripts. The directory tree in Figure \ref{fig:dirtree} can be easily modified and redesigned based on the requirements of the experiments to include markdown notes and analysis files. {For full functionality and documentation for Jobrunner see A. Dhruv \cite{Jobrunner}.}
%
\begin{figure}
\begin{minted}[
    gobble=8,
    frame=single,
  ]{bash}
        # Setting up software and dependencies
        jobrunner setup software/amrex
        jobrunner setup software/flashx

@@

\begin{IEEEbiography}{Akash Dhruv,} is a Postdoctoral Fellow in the Mathematics and Computer Science Division at Argonne National Laboratory where he performs research in DevOps and performance engineering for scientific computing applications, along with building computational pipelines to integrate simulation with machine learning workflows. He received his Ph.D. in mechanical and aerospace engineering from George Washington University, Washington D.C., and a B.Tech in mechanical engineering from National Institute of Technology, Surat. \vspace*{8pt}
\end{IEEEbiography}

\begin{IEEEbiography}{Anshu Dubey,} is a Senior Computational Scientist in the Mathematics and Computer Science Division at Argonne National Laboratory and a Senior Scientist
(CASE) at the University of Chicago. She was previously at Lawrence Berkeley National Laboratory, and before that she was the associate director of the Flash Center for Computational Science at the University of Chicago and also led the CS/Applications
group, who develop, maintain, and distribute the FLASH code. She
received her Ph.D. in computer science from Old Dominion University and
a B.Tech. in electrical engineering from the Indian Institute of Technology,
New Delhi.\vspace*{8pt}
\end{IEEEbiography}
